Fitness tracking reveals task-specific associations between memory, mental health, and exercise

### 2021-09-22 paper/main.tex
finished complete draft of methods section

\section*{Results}
\begin{itemize}
\item characterizing behaviors (potentially broken down by 3--4
  fitness-defined categories, like activity level quartile; may also
  want to have separate figures: one for characterizing behaviors on
  average, and a second breaking down by fitness variable, e.g. after
  the correlation analyses are presented)
  \begin{itemize}
  \item Free recall: pfr, lag-CRP, spc
    \item Naturalistic recall: reproduce a version of the sherlock
      movie/recall trajectories
    \item Foreign language flashcards: p(correct) by activity level
      \item Spatial learning: mean error by number of shapes
      \end{itemize}
    \item Fitness info (break down by task performance, potentiall
      separately for each task); also separate out recent (raw) and recent versus baseline
      \begin{itemize}
      \item activity (steps, zone minutes, floors/elevation)
      \item resting heart rate
        \item sleep
        \end{itemize}
  \item exploratory analysis (correlations)
    \begin{itemize}
    \item Memory-memory
    \item fitness-fitness
    \item survey-survey
    \item (fitness + survey)-memory
    \end{itemize}
  \item predictive analysis (regressions)
    \begin{itemize}
    \item Predict memory performance on held-out task from other tasks
    \item Predict memory performance on each task using fitness data
      \item Predict memory performance on each task using survey data
      \end{itemize}
    \item Reverse correlations: look at recent changes versus baseline trends
      \begin{itemize}
      \item Fitness profile that predicts performance on each task (barplots + timelines)
      \item Fitness profile for each survey demographic (barplots + timelines)
        \begin{itemize}
          \item Select out mental health demographics (based on meds, stress levels)
          \end{itemize}
        \end{itemize}
  \end{itemize}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/experiment}
\caption{\textbf{Battery of memory tasks.}  \textbf{a.  Free recall.}
Participants study 16 words (presented one at a time), followed by an
immediate memory test where they type
each word they remember from the just-studied list.  In the delayed
memory test, participants type any words they remember studying, from
any list.  \textbf{b. Naturalistic recall.}  Participants watch a
brief video, followed by two immediate memory tests.  The first test
asks participants to write out what happened in the video.  The second
test has participants answer a series of multiple choice questions
about the conceptual content of the video.  In the delayed memory
test, participants (again) write out what happened in the video.
\textbf{c. Foreign language flashcards.}  Participants study a
sequence of 10 English-Gailic word pairs, each presented with an
illustration of the given word.  During an immediate memory test,
participants perform a multiple choice test where they select the
Gaelic word that corresponds to the given photograph.  During the
delayed memory test, participants perform a second multiple choice
test, where they select the Gaelic word that corresponds to each of a
new set of photographs.  \textbf{d. Spatial learning.}  In each trial,
participants
study a set of randomly positioned shapes.  Next, the shapes'
positions are altered, and participants are asked to drag the shapes
back to their previous positions.  \textbf{All panels.}  The gray
numbers denote the order in which participants experienced each task
or test.}
\label{fig:tasks}
\end{figure}

@@

Recalled words within an edit distance of 2 (i.e., a Levenshtein Distance less
than or equal to 2) of any word in the wordpool were ``autocorrected''
to their nearest match.  We also manually corrected clear typos or
mispellings by hand (e.g., we corrected ``hippoptumas'' to
``hippopotamus'', ``zucinni'' to ``zucchini'', and so on).  Finally,
we lemmatized each submitted word to match the plurality of the
matching wordpool word (e.g., ``bongo'' was corrected to ``bongos'',
and so on).  After applying these corrections, any submitted words
that matched words presented on the just-studied list were tagged as
``correct'' recalls, and any non-matching words were discarded as
``errors.''  Because participants were not allowed to edit the text
they entered, we chose not to analyze these putative ``errors,'' since
we could not distinguish typos from true misrememberings.

@@

\paragraph{Weight.}  If available, we extracted any weight-related
information affiliated with particiants' Fitbit accounts within 1 year
prior to enrolling in our study.  Depending on their specific device model(s) and
settings, this included their weight, body mass index, and/or body fat
percentage.

\paragraph{Nutrition.} If available, we extracted any nutrition-related
information affiliated with particiants' Fitbit accounts within 1 year
prior to enrolling in our study. Depending on their specific account
settings and usage behaviors, this
included a log of the specific foods they had eaten (and logged)
over the past year, and the amount of water consumed each day.

@@

\paragraph{Activity metrics.}  If available, we extracted any
activity-related information affiliated with particiants' Fitbit
accounts within 1 year prior to enrolling in our study.  Depending on
their specific device model(s) and settings, this included: daily step
counts; daily amount of time spent in each activity level (sedentary,
lightly active, fairly active, or very active, as defined by their
account settings and preferences); daily number of floors climbed;
daily elevation change; and daily total distance traveled.

\subsubsection*{Comparing recent versus baseline measurements.}
We were interested in separating out potential associations between
\textit{absolute} fitness metrics and \textit{relative} metrics.  To
this end, in addition to assessing potential raw
(absolute) fitness metrics, we also defined a simple measure of recent
changes in those metrics, relative to a baseline:
\[
  \Delta_{R, B} m = \frac{\frac{1}{R} \sum_{i = 1}^R
  m(i)}{\frac{1}{B} \sum_{i=R + 1}^{R+B}m(i)},
\]
where $m(i)$ is the value of metric $m$ from $i - 1$ days prior to testing (e.g.,
$m(1)$ represents the value of $m$ on the day the participant accepted
the HIT, and $m(10)$ represents the value of $m$ 9 days prior to
accepting the HIT.  Unless otherwise noted, we set $R = 7$ and $B =
30$.  In other words, to estimate recent changes in any metric $m$, we
divided the average value of $m$ taken over the prior week by the
average value of $m$ taken over the 30 days before that.

\subsubsection*{Exploratory correlation analyses}
We used a bootsrap procedure to identify reliable correlations between
different memory-related, fitness-related, and demographic-related
variables.  For each of $N = 1000$ iterations, we selected (with
replacement) a sample of 113 participants to include.  This yielded,
for each iteration, a sampled ``data matrix'' with one row per sampled
participant and one column for each measured variable.  When
participants were sampled multiple times in a given iteration, as was
often the case, this matrix contained duplicate rows.  We used a
round-robin imputation procedure to estimate the values of any missing
features~\citep{Buck60}.  Next, we computed the Pearson's correlation
between each pair of columns.  This yielded, for each pair of columns,
a distribution of $N$ bootstrapped correlation coefficients.  If fewer
than $95\%$ of the coefficients for a given pair of columns had the
same sign, we excluded the pair from further analysis and considered
the expected correlation between those columns to be undefined.  If
$\geq 95\%$ of the coefficients for a given pair of columns had the
same sign, we computed the expected correlation coefficient as:
\[
  \mathbb{E}_{i, j}\left[ r\right] = \tanh\left(\frac{1}{N} \sum_{n=1}^N
  \tanh^{-1}(\mathrm{corr}\left(m(i)_n, m(j)_n\right))\right),
\]
where $m(x)_n$ represents column $x$ of the bootstrapped data matrix
for iteration $n$, $\tanh$ is the hyperbolic tangent, and $\tanh^{-1}$
is the inverse hyperbolic tangent.

@@

\subsubsection*{Reverse correlation analyses}

>>>>

\section*{Results}
\begin{itemize}
\item characterizing behaviors (potentially broken down by 3--4
  fitness-defined categories, like activity level quartile; may also
  want to have separate figures: one for characterizing behaviors on
  average, and a second breaking down by fitness variable, e.g. after
  the correlation analyses are presented)
  \begin{itemize}
  \item Free recall: pfr, lag-CRP, spc
    \item Naturalistic recall: reproduce a version of the sherlock
      movie/recall trajectories
    \item Foreign language flashcards: p(correct) by activity level
      \item Spatial learning: mean error by number of shapes
      \end{itemize}
    \item Fitness info (break down by task performance, potentially
      separately for each task); also separate out recent (raw) and recent versus baseline
      \begin{itemize}
      \item activity (steps, zone minutes, floors/elevation)
      \item resting heart rate
        \item sleep
        \end{itemize}
  \item exploratory analysis (correlations)
    \begin{itemize}
    \item Memory-memory
    \item fitness-fitness
    \item survey-survey
    \item (fitness + survey)-memory
    \end{itemize}
  \item predictive analysis (regressions)
    \begin{itemize}
    \item Predict memory performance on held-out task from other tasks
    \item Predict memory performance on each task using fitness data
      \item Predict memory performance on each task using survey data
      \end{itemize}
    \item Reverse correlations: look at recent changes versus baseline trends
      \begin{itemize}
      \item Fitness profile that predicts performance on each task (barplots + timelines)
      \item Fitness profile for each survey demographic (barplots + timelines)
        \begin{itemize}
          \item Select out mental health demographics (based on meds, stress levels)
          \end{itemize}
        \end{itemize}
  \end{itemize}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/experiment}
\caption{\textbf{Battery of memory tasks.}  \textbf{a.  Free recall.}
Participants study 16 words (presented one at a time), followed by an
immediate memory test where they type
each word they remember from the just-studied list.  In the delayed
memory test, participants type any words they remember studying, from
any list.  \textbf{b. Naturalistic recall.}  Participants watch a
brief video, followed by two immediate memory tests.  The first test
asks participants to write out what happened in the video.  The second
test has participants answer a series of multiple choice questions
about the conceptual content of the video.  In the delayed memory
test, participants (again) write out what happened in the video.
\textbf{c. Foreign language flashcards.}  Participants study a
sequence of 10 English-Gaelic word pairs, each presented with an
illustration of the given word.  During an immediate memory test,
participants perform a multiple choice test where they select the
Gaelic word that corresponds to the given photograph.  During the
delayed memory test, participants perform a second multiple choice
test, where they select the Gaelic word that corresponds to each of a
new set of photographs.  \textbf{d. Spatial learning.}  In each trial,
participants
study a set of randomly positioned shapes.  Next, the shapes'
positions are altered, and participants are asked to drag the shapes
back to their previous positions.  \textbf{All panels.}  The gray
numbers denote the order in which participants experienced each task
or test.}
\label{fig:tasks}
\end{figure}

@@

Recalled words within an edit distance of 2 (i.e., a Levenshtein Distance less
than or equal to 2) of any word in the wordpool were ``autocorrected''
to their nearest match.  We also manually corrected clear typos or
misspellings by hand (e.g., we corrected ``hippoptumas'' to
``hippopotamus'', ``zucinni'' to ``zucchini'', and so on).  Finally,
we lemmatized each submitted word to match the plurality of the
matching wordpool word (e.g., ``bongo'' was corrected to ``bongos'',
and so on).  After applying these corrections, any submitted words
that matched words presented on the just-studied list were tagged as
``correct'' recalls, and any non-matching words were discarded as
``errors.''  Because participants were not allowed to edit the text
they entered, we chose not to analyze these putative ``errors,'' since
we could not distinguish typos from true misrememberings.

@@

\paragraph{Weight.}  If available, we extracted any weight-related
information affiliated with participants' Fitbit accounts within 1 year
prior to enrolling in our study.  Depending on their specific device model(s) and
settings, this included their weight, body mass index, and/or body fat
percentage.

\paragraph{Nutrition.} If available, we extracted any nutrition-related
information affiliated with participants' Fitbit accounts within 1 year
prior to enrolling in our study. Depending on their specific account
settings and usage behaviors, this
included a log of the specific foods they had eaten (and logged)
over the past year, and the amount of water consumed each day.

@@

\paragraph{Activity metrics.}  If available, we extracted any
activity-related information affiliated with participants' Fitbit
accounts within 1 year prior to enrolling in our study.  Depending on
their specific device model(s) and settings, this included: daily step
counts; daily amount of time spent in each activity level (sedentary,
lightly active, fairly active, or very active, as defined by their
account settings and preferences); daily number of floors climbed;
daily elevation change; and daily total distance traveled.

\subsubsection*{Comparing recent versus baseline measurements.}
We were interested in separating out potential associations between
\textit{absolute} fitness metrics and \textit{relative} metrics.  To
this end, in addition to assessing potential raw
(absolute) fitness metrics, we also defined a simple measure of recent
changes in those metrics, relative to a baseline:
\[
  \Delta_{R, B} m = \frac{B \sum_{i = 1}^R
  m(i)}{R \sum_{i=R + 1}^{R+B}m(i)},
\]
where $m(i)$ is the value of metric $m$ from $i - 1$ days prior to testing (e.g.,
$m(1)$ represents the value of $m$ on the day the participant accepted
the HIT, and $m(10)$ represents the value of $m$ 9 days prior to
accepting the HIT.  Unless otherwise noted, we set $R = 7$ and $B =
30$.  In other words, to estimate recent changes in any metric $m$, we
divided the average value of $m$ taken over the prior week by the
average value of $m$ taken over the 30 days before that.

\subsubsection*{Exploratory correlation analyses}
We used a bootstrap procedure to identify reliable correlations between
different memory-related, fitness-related, and demographic-related
variables.  For each of $N = 1000$ iterations, we selected (with
replacement) a sample of 113 participants to include.  This yielded,
for each iteration, a sampled ``data matrix'' with one row per sampled
participant and one column for each measured variable.  When
participants were sampled multiple times in a given iteration, as was
often the case, this matrix contained duplicate rows.  We used a
round-robin imputation procedure to estimate the values of any missing
features~\citep{Buck60}.  Next, we computed the Pearson's correlation
between each pair of columns.  This yielded, for each pair of columns,
a distribution of $N$ bootstrapped correlation coefficients.  If fewer
than $95\%$ of the coefficients for a given pair of columns had the
same sign, we excluded the pair from further analysis and considered
the expected correlation between those columns to be undefined.  If
$\geq 95\%$ of the coefficients for a given pair of columns had the
same sign, we computed the expected correlation coefficient as:
\[
  \mathbb{E}_{i, j}\left[ r\right] = \tanh\left(\frac{1}{N} \sum_{n=1}^N
  \tanh^{-1}(\mathrm{corr}\left(m(i)_n, m(j)_n\right))\right),
\]
where $m(x)_n$ represents column $x$ of the bootstrapped data matrix
for iteration $n$, $\tanh$ is the hyperbolic tangent, and $\tanh^{-1}$
is the inverse hyperbolic tangent.

@@

\subsubsection*{Reverse correlation analyses}
We sought to characterize potential associations between the history
of participants' fitness-related activities leading up to the time
they participated in a memory task and their performance on the given
task.  For each fitness-related variable, we constructed a timeseries
matrix whose rows corresponded to timepoints (sampled once per day)
leading up to the day the participant accepted the HIT for our study,
and whose columns corresponded to different participants.  These
matrices often contained missing entries, since different
participants' Fitbit devices tracked fitness-related activities
differently.  For example, participants whose Fitbit devices lacked
heart rate sensors would have missing entries for any heart
rate-related variables.  Or, if a given participant neglected to wear
their fitness tracker on a particular day, the column corresponding to
that participant would have missing entries for that day.

In addition to this set of matrices storing timeseries data for each
fitness-related variable, we also constructed a memory performance
matrix, $M$, whose rows corresponded to different memory-related variables,
and whose columns corresponded to different participants.  For
example, one row of the memory performance matrix reflected the
average proportion of words (across lists) that each
participant remembered during the immediate free recall test, and so
on.

Given a fitness timeseries matrix, $F$, we computed the weighted
average and weighted standard error of the mean of each row of $F$,
where the weights were given by a particular memory-related variable
(row of $M$).  For example, if $F$ contained participants' daily step
counts, we could use any row of $M$ to compute a weighted average
across any participants who contributed step count data on each day.
Choosing a row of $M$ that corresponded to participants' performance
on the naturalistic recall task would mean that participants who
performed better on the naturalistic recall task would contribute more
to the weighted average timeseries of daily step counts.
Specifically, for each row, $t$, of $F$, we computed the weighted
average (across the $S$ participants)  as:
\[
\bar{f}(t) = \sum_{s=1}^S \dot{m}(s) F(t, s),
\]
where $\dot{m}$ denotes the normalized min-max scaling of $m$ (the row of $M$
corresponding to the chosen memory-related variable):
\[
  \dot{m} = \frac{m}{\sum_{s=1}^S \hat{m}(s)},
\]
where
\[
  \hat{m} = \frac{m - \min(m)}{\max(m) - \min(m)}
\]

We computed the weighted standard error of the mean as:
\[
\mathrm{SEM}_m\left(f(t)\right) = \frac{\left| \sum_{s=1}^S \left( F(t, s) -
    \bar{f}(t)\right) \right|}{\sqrt{S}}.
\]

When a given row of $F$ was missing data from one or more
participants, those participants were excluded from the weighted
average for the corresponding timepoint and the weights (across all
remaining participants) were re-normalized to sum to 1.
The above procedure yielded, for each memory
variable, a timeseries of average (and standard error of the mean)
fitness tracking values leading up to the day of the experiment.
